---
title: "Gauge-calibrated proportion mediated: the residual W"
format: html
vignette: >
  %\VignetteIndexEntry{Gauge-calibrated proportion mediated: the residual W}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2026)
# Quarto renders vignettes in a child process whose library path may not
# include the just-built package during R CMD check (notably on Windows).
# Degrade to code-only there instead of erroring on library(probmed); Linux/
# macOS builds (CRAN, r-universe) still execute with full output.
if (!requireNamespace("probmed", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

## The problem with a single proportion-mediated number

The interventional proportion mediated `P_med = IIE / OE` summarizes mediation
in one number. That summary is only trustworthy when the overall effect
**decomposes additively** into a direct and an indirect part. With a
treatment-by-mediator interaction (or intermediate confounding) the additive
split fails: there is a leftover

$$
R = \mathrm{OE} - \mathrm{IDE} - \mathrm{IIE},
$$

the **non-decomposability remainder**. When $R$ is nonzero, `IIE / OE` is reporting a
fraction of a quantity that was never the sum of its parts.

`ward_residual()` estimates `P_med` *and* the **gauge residual**

$$
W = R / \mathrm{OE},
$$

a scale-free flag for that failure. $W \approx 0$ certifies the additive split;
a $W$ significantly away from 0 says "interpret `P_med` with care."

```{r}
#| label: load
library(probmed)
```

## A clean (additive) example

With no interaction term, $R$ — and hence $W$ — should sit at zero, and the
gauge `P_med` agrees with the usual story.

```{r}
#| label: additive
n <- 1500
C <- rnorm(n)
A <- rbinom(n, 1, plogis(-0.2 + 0.8 * C))
M <- 0.6 * A + 0.4 * C + rnorm(n)
Y <- 0.5 * A + 0.7 * M + 0.3 * C + rnorm(n)        # no A*M term
ward_residual(data.frame(A, M, Y, C))
```

The `W` confidence interval covers 0: the single-number `P_med` is safe here.

## An interaction breaks the split

Add an `A * M` term and the gauge lights up — `W` is now significantly positive
and the printout warns that the additive split is unreliable.

```{r}
#| label: interaction
Yi <- 0.5 * A + 0.7 * M + 0.8 * A * M + 0.3 * C + rnorm(n)
ward_residual(data.frame(A, M, Y = Yi, C))
```

## Non-binary-coded exposures: `a0` / `a1`

The treatment need not be coded `0/1`. Any **two-level** coding works — a
factor, `{1, 2}`, `{-1, 1}` — by naming the reference `a0` and comparison `a1`.
`A` is recoded internally to the indicator `(A == a1)`.

```{r}
#| label: factor-coding
dat <- data.frame(A = factor(ifelse(A == 1, "drug", "placebo")),
                  M = M, Y = Yi, C = C)
ward_residual(dat, a0 = "placebo", a1 = "drug")
```

`W = R / OE` is **anti-symmetric** in the contrast direction: swapping `a0` and
`a1` flips the sign of `W` (the remainder `R` is invariant, `OE` changes sign)
while leaving its magnitude — the thing you interpret — unchanged.

A treatment with **more than two** levels is rejected rather than silently
subsetted: restricting the data to two of several levels would change the
population the corner means are averaged over, estimating a different
(sub-population) gauge. Filter to the two levels you intend to contrast first.

```{r}
#| label: multi-level-error
#| error: true
multi <- data.frame(A = sample(0:2, n, replace = TRUE), M = M, Y = Yi, C = C)
ward_residual(multi, a0 = 0, a1 = 1)
```

## Inference: what the two arms actually do

`W` and `P_med` are **ratios**. The default `se_method = "analytic"` reports the
influence-function standard error of the cross-fit one-step estimator and a
symmetric Wald interval. Measured against exact truth
(`inst/sim/se_shipped_check.R`, 250 datasets per cell, n = 800) it covers 0.936
for a continuous outcome and 0.972 for a binary one under strong identification,
0.912 when the total effect is moderate, and its dispersion from dataset to
dataset matches that of an oracle that knows the nuisance functions. `reps > 1`
averages the influence matrix over several independent partitions; it removes
the fold-split component of the variance, which with correct weights is small
(empSD 0.063 at `reps = 1` vs 0.061 at `reps = 10`, oracle 0.056):

```{r}
#| label: reps
ward_residual(data.frame(A, M, Y = Yi, C), reps = 10)
```

A note on history, because the earlier documentation said the opposite: before
2026-08-22 the corner influence function used one inverse-probability weight per
fold — `ifelse()` on a scalar test silently returned the first fold row's
propensity for every row — and every coverage figure the package reported
(Wald ~0.85–0.90, percentile ~1.00, the gate-validation grid, the variance
decomposition in `docs/specs/FINDINGS-2026-08-22-phi-decomposition.md`) was
produced with that bug. Those numbers are superseded; the bootstrap arm and the
two gates were re-measured the same day
(`docs/specs/FINDINGS-2026-08-22-postfix-bootstrap-gates.md`).

Setting `se_method = "bootstrap"` switches the `W` and `P_med` intervals to
**percentile** intervals of a nonparametric bootstrap that resamples rows and
refits the whole cross-fit, with a fresh partition each resample. With correct
weights it buys nothing: percentile coverage 0.905–0.930 in the regular cells at
the same width as the Wald interval, and a bootstrap standard error that is
unstable for binary outcomes. It remains available for users who want a refit
bootstrap. The `weak_id` flag it enables — percentile interval at least 3× wider
than Wald — is a symptom of bootstrap instability near `OE = 0`, not a coverage
diagnostic: flagged and unflagged draws cover alike at every threshold tried; it
no longer warns. The printed label tracks which construction produced the
interval.

```{r}
#| label: bootstrap
ward_residual(data.frame(A, M, Y = Yi, C), se_method = "bootstrap", B = 200)
```

For the near-null / weak-identification regime — `OE` not bounded away from 0 —
the ratio itself is non-regular: in the grid's near-null cells even an oracle
with known nuisances produces exploding standard errors, and `oe_regular = FALSE`
flags exactly this. Prefer the **Fieller** confidence set there
(`fieller = TRUE`, on by default), which is honest about being unbounded when the
total effect is not significant, and report the unnormalized effects.

## Summary

- `ward_residual()` reports `P_med` together with the gauge residual
  `W = R/OE`; `W` flags when the additive direct/indirect split — and therefore
  the single-number `P_med` — is unreliable.
- Use `a0` / `a1` for any two-level exposure coding; `>2` levels is an error by
  design.
- For ratio inference, the default analytic Wald interval is calibrated under
  regular identification (0.936 continuous / 0.972 binary at n = 800);
  `reps > 1` adds a small efficiency gain; `se_method = "bootstrap"` gives
  percentile intervals that are no better (0.905–0.930) and its `weak_id` flag
  is not a coverage diagnostic; `oe_regular` and the Fieller set handle the
  near-null case, where the ratio is non-regular.
