---
title: "Introduction to probmed"
format: html
vignette: >
  %\VignetteIndexEntry{Introduction to probmed}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

The `probmed` package provides a framework for computing **$P_{med}$**, a scale-free probabilistic effect size for causal mediation analysis. Unlike traditional effect sizes like the "Proportion Mediated" ($P_M$), which can be unstable or difficult to interpret (especially with inconsistent mediation or non-linear models), $P_{med}$ offers a clear probabilistic interpretation.

### What is $P_{med}$?

$P_{med}$ is defined as the probability that the outcome for an individual in the treatment group ($Y(1)$) is superior to the outcome for an individual in the control group ($Y(0)$), specifically due to the indirect effect transmitted through the mediator ($M$).

Mathematically, for a treatment $X$, mediator $M$, and outcome $Y$:

$$ P_{med} = P(Y(1, M(1)) > Y(0, M(0))) $$

where the direct effect of $X$ on $Y$ is held constant or accounted for, focusing purely on the mediation path.

Key advantages:

1.  **Scale-Free**: It does not depend on the units of $Y$ or $M$.
2.  **Bounded**: It ranges from 0 to 1 (or 0.5 for no effect in some contexts).
3.  **Robust**: It works well for both linear and generalized linear models (GLMs).

## Installation

You can install the development version of `probmed` from GitHub:

```r
# install.packages("devtools")
devtools::install_github("data-wise/probmed")
```

## Basic Example: Linear Mediation

Let's demonstrate `probmed` with a simple linear mediation model. We will simulate data where $X$ affects $M$, and both $X$ and $M$ affect $Y$.

### 1. Simulate Data

```{r}
#| label: setup
library(probmed)

set.seed(123)
n <- 500

# Confounder / Covariate
C <- rnorm(n)

# Treatment (Continuous)
X <- rnorm(n)

# Mediator: M ~ X + C
# Path a = 0.5
M <- 0.5 * X + 0.3 * C + rnorm(n)

# Outcome: Y ~ M + X + C
# Path b = 0.4, Path c' = 0.2
Y <- 0.4 * M + 0.2 * X + 0.2 * C + rnorm(n)

data <- data.frame(X, M, Y, C)
head(data)
```

### 2. Estimate $P_{med}$

To estimate $P_{med}$, we use the `pmed()` function. We need to specify:

*   The **outcome model** formula (`Y ~ ...`).
*   The **mediator model** formula (`formula_m = ...`).
*   The names of the **treatment** and **mediator** variables.

We will use the **parametric bootstrap** method for inference, which is fast and accurate for standard models.

```{r}
#| label: estimate
result <- pmed(
    Y ~ X + M + C, # Outcome model
    formula_m = M ~ X + C, # Mediator model
    data = data,
    treatment = "X",
    mediator = "M",
    method = "parametric_bootstrap",
    n_boot = 1000, # Number of bootstrap samples
    seed = 42
)
```

### 3. Interpret Results

We can print and summarize the results.

```{r}
#| label: print
print(result)
```

The output shows the estimated $P_{med}$ and its 95% confidence interval.

*   **Estimate**: A value significantly greater than 0.50 indicates a positive mediation effect.
*   **Confidence Interval**: If the interval excludes 0.50, the effect is statistically significant.

You can also get a more detailed summary:

```{r}
#| label: summary
summary(result)
```

## Advanced Example: Binary Outcome (GLM)

`probmed` shines when dealing with non-linear models where traditional coefficients ($a \times b$) are hard to interpret. Let's look at a case with a **binary outcome**.

### 1. Simulate Binary Data

```{r}
#| label: sim-binary
set.seed(456)
n <- 500
X <- rnorm(n)
M <- 0.5 * X + rnorm(n)

# Outcome is binary (Logistic model)
# logit(P(Y=1)) = 0.5*M + 0.2*X
logits <- 0.5 * M + 0.2 * X
probs <- 1 / (1 + exp(-logits))
Y_bin <- rbinom(n, 1, probs)

data_bin <- data.frame(X, M, Y_bin)
```

### 2. Estimate with GLM

We simply specify `family_y = binomial()` to tell `probmed` that the outcome model is a logistic regression.

```{r}
#| label: estimate-binary
result_bin <- pmed(
    Y_bin ~ X + M,
    formula_m = M ~ X,
    data = data_bin,
    family_y = binomial(), # Specify binomial family
    treatment = "X",
    mediator = "M",
    method = "parametric_bootstrap",
    n_boot = 1000,
    seed = 42
)

print(result_bin)
```

The interpretation remains the same: $P_{med}$ is the probability of the outcome being "better" (higher latent utility or probability class) due to mediation.

## Conclusion

The `probmed` package makes it easy to compute interpretable, scale-free effect sizes for mediation analysis. Whether you are working with simple linear models or complex GLMs, $P_{med}$ provides a unified metric for understanding the magnitude of indirect effects.
