---
title: "Parallel mediation: joint P_med"
format: html
vignette: >
  %\VignetteIndexEntry{Parallel mediation: joint P_med}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2026)
# Quarto renders vignettes in a child process whose library path may not
# include the just-built package during R CMD check (notably on Windows).
# Degrade to code-only there instead of erroring on library(probmed); Linux/
# macOS builds (CRAN, r-universe) still execute with full output.
if (!requireNamespace("probmed", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

## Overview

When a treatment influences an outcome through **several mediators at once** —
each modelled in its own equation but all entering a single outcome equation —
the structure is *parallel mediation*:

$$
X \rightarrow M_1, \quad X \rightarrow M_2, \quad \dots, \quad X \rightarrow M_k,
\qquad Y \sim X + M_1 + \dots + M_k .
$$

`probmed` extends $P_{med}$ to this setting with a **joint** estimand: the
probability that the outcome is higher when *all* mediators are set to their
treated levels than when all are set to their control levels. This vignette
derives the estimand, gives its closed form, and demonstrates the four inference
methods. It covers the **Gaussian** case (continuous mediators and outcome).

## The joint estimand

Write the treated and control treatment levels as $x = $ `x_value` and
$x^\ast = $ `x_ref`, with contrast $\delta = x - x^\ast$. As in the
single-mediator case, the treatment is **held at $x$ for both** potential
outcomes; only the mediators vary between their treated levels $M_j(x)$ and
control levels $M_j(x^\ast)$, drawn independently. The direct effect $c'$
therefore cancels and $P_{med}$ reflects mediation only.

Each mediator follows $M_j = i_{m_j} + a_j X + \varepsilon_{m_j}$ with
$\varepsilon_{m_j} \sim N(0, \sigma_{m_j}^2)$, and the outcome follows
$Y = i_y + c' X + \sum_j b_j M_j + \varepsilon_y$ with
$\varepsilon_y \sim N(0, \sigma_Y^2)$. The two potential outcomes (treatment
fixed at $x$) are

$$
Y_t = i_y + c' x + \sum_j b_j M_j(x) + \varepsilon_t,
\qquad
Y_c = i_y + c' x + \sum_j b_j M_j(x^\ast) + \varepsilon_c .
$$

Each mediator difference is Gaussian and independent across $j$,

$$
M_j(x) - M_j(x^\ast) \sim N\!\big(a_j\delta,\; 2\sigma_{m_j}^2\big),
$$

so the outcome difference is Gaussian with

$$
Y_t - Y_c \;\sim\; N\!\Big(\; \delta \textstyle\sum_j a_j b_j,\;\;
2\textstyle\sum_j b_j^2 \sigma_{m_j}^2 + 2\sigma_Y^2 \;\Big).
$$

The **joint parallel $P_{med}$** is then a clean closed form:

$$
\boxed{\;P_{med} = \Phi\!\left(
\frac{\delta \sum_j a_j b_j}
     {\sqrt{2\sum_j b_j^2 \sigma_{m_j}^2 + 2\sigma_Y^2}}
\right)\;}
$$

which recovers the single-mediator formula
$\Phi\!\big(ab / \sqrt{2(b^2\sigma_M^2 + \sigma_Y^2)}\big)$ exactly at $k = 1$.
Reversing the contrast ($\delta \to -\delta$) maps
$P_{med} \to 1 - P_{med}$. The associated **total indirect effect** is
$\sum_j a_j b_j$.

## A worked example

We simulate two parallel mediators, fit each equation, and extract a
`ParallelMediationData` object with **medfit**.

```{r}
#| label: data
n <- 1500
X <- rnorm(n)
M1 <- 0.5 * X + rnorm(n)
M2 <- 0.4 * X + rnorm(n)
Y <- 0.3 * X + 0.5 * M1 + 0.6 * M2 + rnorm(n)
dat <- data.frame(X, M1, M2, Y)

fit_m1 <- lm(M1 ~ X, data = dat)
fit_m2 <- lm(M2 ~ X, data = dat)
fit_y <- lm(Y ~ X + M1 + M2, data = dat)

ex <- medfit::extract_mediation(
  object = fit_m1, model_y = fit_y,
  treatment = "X", mediator = c("M1", "M2"),
  mediator_models = list(fit_m2),
  structure = "parallel", data = dat
)
```

### Plugin (point estimate)

```{r}
#| label: plugin
library(probmed)
pmed(ex, method = "plugin")
```

The point estimate is the closed form above; the indirect effect is
$\sum_j a_j b_j$.

### Parametric bootstrap

Draws the structural coefficients $(a_1, \dots, b_k)$ from
$N(\hat\theta, \widehat{\mathrm{Var}})$ (residual SDs held fixed) and evaluates
the closed form per draw.

```{r}
#| label: parametric
pmed(ex, method = "parametric_bootstrap", n_boot = 500, seed = 1)
```

### Nonparametric bootstrap

Resamples rows, refits the mediator and outcome models, and evaluates the closed
form per resample.

```{r}
#| label: nonparametric
pmed(ex, method = "nonparametric_bootstrap", n_boot = 300, seed = 1)
```

### MBCO (deterministic)

Inverts a constrained likelihood-ratio test for both the joint $P_{med}$ and the
total indirect effect — no resampling, so it is seed-free and
grid-resolution-independent.

```{r}
#| label: mbco
pmed(ex, method = "mbco")
```

## Scope and limitations

- **Gaussian only.** Parallel $P_{med}$ requires continuous mediators and a
  continuous outcome. A non-Gaussian `ParallelMediationData` (one lacking finite
  residual SDs) raises an informative error; non-Gaussian support is blocked
  pending family slots in the medfit parallel container.
- **Joint estimand only.** This release reports the single joint $P_{med}$ for
  moving all mediators together. Path-specific ($P_{med}$ per mediator)
  decomposition is planned future work.
- **Object entry.** Reach the method via
  `medfit::extract_mediation(..., structure = "parallel")` then `pmed()`; a
  formula convenience wrapper is planned.
- **Serial / sequential** mediation ($X \to M_1 \to M_2 \to Y$) is a separate
  estimand and is not covered here.
